\large{\bf Large effects and the infinitesimal model }

### 2023-07-05 writeup.tex
citations

\comment{TODO: citations below}

As suggested by its name,
under the infinitesimal model the effects of any given genetic variant on the trait
is small (indeed, vanishing),
and selection-driven trait shifts are mediated by small changes in frequencies of many alleles of small effect.
However, empirical understanding of the genetic basis of trait variation in real-world species
often finds a significant portion of heritable variation explained by variants at only a few loci.
There are many possible explanations for this observation,
including selection or publication bias,
filters for larger-effect variants due to spatially varying selection (cite),
and/or grouping together of concordant alleles by linkage or inversions (cite).
However, it seems clear that in at least those cases where we understand the underlying molecular mechanisms,
mutations of large effect can be important in practice
(cite examples: insecticide resistance, coloration, etcetera).
The consensus in the field seems to be that traits exist on a spectrum between monogenic and polygenic,
and that many - if not most - are firmly in the intermediate, ``oligogenic'' region (cite),
where heritable trait variation is due both to segregating large- and small-effect variants.
Indeed, some sensible verbal models of the biological mechanisms by which a genetic change
comes to have an effect on an organism's trait
describe a hierarchy or range of ``proximities'' to the trait,
by which changes to more ``nearby'' genes have the possibility of causing larger changes in the trait
(cite omnigenic and others).
This suggests a mixture model for the effect size distribution:
perhaps the distribution of effect sizes within each gene is Normal,
but with a standard deviation that varies with ``proximity'' to the trait;
such mixtures can easily have large proportions of extreme values.

The Gaussian distribution appears thanks to ``the'' Central Limit Theorem,
i.e., the universality
of adding up a great many very small and independent things.
However, there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws}.
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \}$ is proportional to $t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, an $\alpha$-stable distribution
(and the approximation becomes exact, as $n \to \infty$).
In such ensembles, single large entries can be important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.

@@

Stable distributions are by no means new to population biology.
For instance, it is a long-standing question how often the motions of organisms
are better modeled by essentially Brownian random walks or by L\'evy processes (cite),
and a growing literature studies the effects of such power law distributions
on the shapes and dynamics of traveling waves such as biological invasions (cite).
\comment{Other examples?}
More directly analogous to the problem we study here, \citet{landis2012phylogenetic}
found that an $\alpha$-stable process (with $\alpha \approx 1.5$) fit the evolution of log endocranial volume
on the primate phylogeny
better than either a Brownian model.
It is also important to note that
the infinitesimal model implies a Gaussian distribution for the difference between an offspring's trait and their parents' average,
\emph{not} for the entire population.
The question of when the trait distribution in an evolving population
might be modeled as a Gaussian has been discussed extensively, e.g., by \comment{cite}.

The ways in which genetic variants affect an organisms' trait can be complex,
and interactions between alleles, between loci, and with the environment are often important in practice.
In fact, many traits claimed to be genetically determined
in fact only appear heritable through non-genetic means,
e.g., systematic inequalities in resource distribution partitioned by class or race. \comment{e.g., cite}
In this paper, we follow the literature in working with the relatively simple purely additive case
of a trait that can be decomposed into a sum across loci of the contributions of the alleles from each locus,
even omitting an independent effect of ``the environment''.
However, we emphasize that such models should \emph{not} always be the default for understanding biological variation,
especially for human traits.

>>>>

As suggested by its name,
under the infinitesimal model the effects of any given genetic variant on the trait
is small (indeed, vanishing),
and selection-driven trait shifts are mediated by small changes in frequencies of many alleles of small effect.
However, empirical understanding of the genetic basis of trait variation in real-world species
often finds a significant portion of heritable variation explained by variants at only a few loci.
There are many possible explanations for this observation,
including selection or publication bias,
filters for larger-effect variants due to spatially varying selection
\citep{barton1987establishment,yeaman2011establishment,yeaman2011genetic},
and/or grouping together of concordant alleles by linkage or inversions
\citep{kirkpatrick2006chromosome}.
However,
it seems clear that in at least those cases where we understand the underlying molecular mechanisms,
mutations of large effect can be important in practice.
% \citep[e.g.,][]{cyclodiene-resistance,hoekstra2006genetics}.
The consensus in the field seems to be that traits exist on a spectrum between monogenic and polygenic,
and that many - if not most - lie in the intermediate region,
where heritable trait variation is due both to segregating large- and small-effect variants.
Indeed, some sensible verbal models of the biological mechanisms by which a genetic change
comes to have an effect on an organism's trait
describe a hierarchy or range of ``proximities'' to the trait,
by which changes to more ``nearby'' genes have the possibility of causing larger changes in the trait
\citep{kopp2012transcriptional,boyle2017expanded}.
This suggests a mixture model for the effect size distribution:
perhaps the distribution of effect sizes within each gene is Normal,
but with a standard deviation that varies with ``proximity'' to the trait;
such mixtures can easily have large proportions of extreme values.

The Gaussian distribution appears thanks to ``the'' Central Limit Theorem,
i.e., the universality of adding up a great many very small and independent things.
However, there is a wider class of Central Limit Theorems,
which describe the distributions of sums of a great many independent and exchangeable things:
the \emph{stable laws} \citep{gnedenko1968limit}
Roughly speaking, these say that if we add together $n$ independent copies of a random variable $X$
for which $\P\{ X > t \}$ is proportional to $t^{-\alpha}$ for some $0 < \alpha < 2$,
then -- almost regardless of the details -- the result, scaled by $n^{1/\alpha}$,
is well-approximated by a universal form, an $\alpha$-stable distribution
(and the approximation becomes exact, as $n \to \infty$).
In such ensembles, single large entries can be important:
concretely, the largest of the $n$ copies will be of order $n^{1/\alpha}$,
so that e.g., for $\alpha=1$ a positive proportion of the sum of $n$ entries will be contributed by a single one.

@@

Stable distributions are by no means new to population biology.
For instance, it is a long-standing question how often the motions of organisms
are better modeled by essentially Brownian random walks or by L\'evy processes \citep{benhamou2007animals}
and a growing literature studies the effects of such power law distributions
on spatial genetic and dynamics
\citep[e.g.,][]{paulose2019spatial,smith2023isolation}.
More directly analogous to the problem we study here, \citet{landis2012phylogenetic}
found that an $\alpha$-stable process (with $\alpha \approx 1.5$) fit the evolution of log endocranial volume
on the primate phylogeny
better than either a Brownian model.
It is also important to note that
the infinitesimal model implies a Gaussian distribution for the difference between an offspring's trait and their parents' average,
\emph{not} for the entire population.
The question of when the trait distribution in an evolving population
might be modeled as a Gaussian has been discussed extensively
\citep[e.g.,][]{turelli1994genetic,dbarre2015evolution}.

The ways in which genetic variants affect an organisms' trait can be complex,
and interactions between alleles, between loci, and with the environment are often important in practice.
% In fact, many traits claimed to be genetically determined
% in fact only appear heritable through non-genetic means,
% e.g., systematic inequalities in resource distribution partitioned by class or race
% \citep[e.g.,][]{davenport1916hereditary}.
In this paper, we follow the literature in working with the relatively simple purely additive case
of a trait that can be decomposed into a sum across loci of the contributions of the alleles from each locus,
even omitting an independent effect of ``the environment''.
However, we emphasize that such models should \emph{not} always be the default for understanding biological variation,
especially for human traits.
